Typed Clojure

### 2013-07-26 schemeworkshop13.tex
Work on judgements

\documentclass[preprint]{sigplanconf}

@@

\title{Typed Clojure}
%\subtitle{Subtitle Text, if any}

\authorinfo{Name1}
           {Affiliation1}
           {Email1}
\authorinfo{Name2\and Name3}
           {Affiliation2/3}
           {Email2/3}

@@

\begin{abstract}
This is the text of the abstract.
\end{abstract}

@@

\keywords
keyword1, keyword2

@@

@@

@@

\begin{figure*}
$$
\begin{altgrammar}
  \expd{}, \expe{} &::=& (\expe{}\ \expe{}) \alt \abs{\x a}{\expe{}}
                &\mbox{Expressions}
%  \c{} &::=& \assoc{} \alt \dissoc{} \alt \get{}
%              &\mbox{Constants}
\end{altgrammar}
$$
\end{figure*}

@@

Armed with general unions and the type \lstinline|nil|, 
we can make Typed Clojure's reference types \emph{non-nullable}.
A Java type like \emph{java.lang.Number} is written \lstinline|(U nil java.lang.Number)|
in Typed Clojure.

@@

A set of conversion rules define how to convert a Java type signature into a null-safe
Clojure version. These rules are designed to be conservative and can be explicitly overridden
by programmers. Method parameters are non-nullable, while return types are nullable.
Field types are nullable. Constructors parameters are non-nullable, and return types
are also non-nullable: it is a Java invariant that all constructors return an instance
of the Class they are constructing.

>>>>

\documentclass[preprint,10pt]{sigplanconf}

@@

\title{Typed Clojure}
\subtitle{A Practical Type System for Clojure based on Typed Racket}

\authorinfo{Ambrose Bonnaire-Sergeant}
           {}
           {abonnairesergeant@gmail.com}
\authorinfo{Rowan Davies}
           {University of Western Australia}
           {rowan@csse.uwa.edu.au}

@@

\begin{abstract}
Typed Racket has proved to be a valuable starting point for
a gradual type system for Clojure.
Building a similar type system for a new language gives the
designer some flexibility to repurpose and extend features.
This paper gives an overview of Typed Clojure, concentrating
on the extensions and differences from Typed Racket. We also
show where Typed Racket's features were particularly useful
for type checking non-trivial Clojure idioms.
\end{abstract}

@@

\keywords
gradual types, Clojure

@@

\section{Contributions}

The main result of this paper is to demonstrate that Typed Racket's design
and implementation is applicable to a practical type system for Clojure
by implementing Typed Clojure.

@@

\section{Multimethods}

Clojure provides multimethods which can dispatch on the result of an
arbitrary function, the \emph{dispatch function}. \emph{Methods} are
added to the multimethod
which are associated with a \emph{dispatch value}, which is compared to the
result of apply the dispatch function to the argument passed to the
multimethod. If this comparison unambiguously dispatches to one method,
that method is then called with the current arguments.

Occurrence typing, in particular when extended with paths,
for checking method definitions for arbitrary dispatch functions.

\begin{lstlisting}
(ns example.mm
  (:require 
    [clojure.core.typed :refer [ann]]))

(ann my-mm [Any -> Number])
(defmulti my-mm class)

(defmethod my-mm Float [f] (inc f))
\end{lstlisting}

\begin{lstlisting}
(ann clojure.core/class
  [Any -> (U nil Class)
   :object {:id 0, :path [Class]}])
\end{lstlisting}

\begin{verbatim}
TODO
- example of normal MM usage
- explain descrepency btwn TC syntax and formal
  TR calculs syntax. eg. :object/:filter vs over arrow
\end{verbatim}

@@

\begin{figure*}
$$
\begin{altgrammar}
  \expd{}, \e{} &::=& \x\ \alt {\comb {\e{}} {\e{}}} %\alt {\abs {\x a} {\e{}}}
                      \alt \true{} \alt \false{} \alt \nil{}
                      \alt {\tryexp {\e{}} {\class} {\x} {\e{}} {\e{}}}
                      \alt {\ifexp {\e{}} {\e{}} {\e{}}}
                      \alt {\doexp {\e{}} {\e{}}}
                      \alt {\fnexp {\x{}} {\t{}} {\e{}}}
                &\mbox{Expressions} \\ \\
  \s{}, \t{}    &::=& (\class{}\ \overrightarrow{\t{}})
                      \alt (\ma{\Value\ \v})
                      \alt (\cup\ \overrightarrow{\t{}})
                      \alt (\cap\ \overrightarrow{\t{}})
                      \alt (\Not\ \t{})
                      \alt (\All\ [\overrightarrow{\x{}}] \t{})
                      \alt (\TFn\ [\overrightarrow{\x{}}] \t{})
                &\mbox{Types}
%  \c{} &::=& \assoc{} \alt \dissoc{} \alt \get{}
%              &\mbox{Constants}
\end{altgrammar}
$$
\caption{Syntax of Expressions, Types}
\end{figure*}

\begin{figure*}
\begin{mathpar}
\infer [T-Var]
{ \inpropenv {\propenv{}} {\isprop {\x} {\tau}}}
{ \judgement {\propenv{}} {{\hastype {\x{}} {\t{}}}} {{\filterset {{\notprop {\falsy} {\x{}}}} {{\isprop {\falsy} {\x{}}}}}}
             {\emptyobject{}} {\topprop{}}}

\infer [T-True]
{}
{ \judgement {\propenv{}} {{\hastype {\true{}} {\True{}}}} {{\filterset {\topprop{}} {\botprop{}}}} {\emptyobject{}} {\topprop{}}}

\infer [T-False]
{}
{ \judgement {\propenv{}} {{\hastype {\false{}} {\False{}}}} {{\filterset {\botprop{}} {\topprop{}}}} {\emptyobject{}} {\topprop{}}}

\infer [T-Nil]
{}
{ \judgement {\propenv{}} {{\hastype {\nil{}} {\Nil{}}}} {{\filterset {\botprop{}} {\topprop{}}}} {\emptyobject{}} {\topprop{}}}

\infer [T-Do]
{ \judgement {\propenv{}} {{\hastype {\e1} {\t1}}} {{\filterset {\prop{1+}} {\prop2}}} {\object{1}} {{\prop{f1}}}
\\
  \judgement {\propenv{}, \prop{f1}} {{\hastype {\e2} {\t2}}} {{\filterset {\prop3} {\prop4}}} {\object{}} {\prop{f2}}
  }
{ \judgement {\propenv{}} {{\hastype {\doexp {\e1} {\e2}} {\t2}}} {{\filterset {\prop3} {\prop4}}} {\object} 
             {{\andprop {\prop{f1}} {\prop{f2}}}}}

\infer [T-If]

%\infer [T-App]

\end{mathpar}
\caption{Typing Rules}
\end{figure*}

@@

Armed with general unions and the type \lstinline|nil|, 
we make Typed Clojure's reference types \emph{non-nullable}.
A Java type like \emph{java.lang.Number} is written \lstinline|(U nil java.lang.Number)|
in Typed Clojure.

@@

A set of conversion rules define how to convert a Java type signature into a null-safe
Clojure version. These rules are designed to be conservative and can be explicitly overridden
by programmers. Method parameters are non-nullable, while return types are nullable.
Field types are nullable. Constructor parameters are non-nullable, and return types
are also non-nullable: it is a Java invariant that all constructors return an instance
of the class they are constructing.

\begin{verbatim}
TODO 
- write conversion rules for 
  Java type -> Clojure type
\end{verbatim}
